---
title: "Introduction to lpSolveR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to lpSolveR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
library(lpSolveR)
```

## 1. Introduction

Linear programming (LP) is one of the most widely used tools in operations
research, logistics, economics, and data science.  **lpSolveR** is an R package
that provides pure-R implementations of three classical LP algorithms:

* The **simplex method** — a pivot-based algorithm that traverses the vertices
  of the feasible polytope.
* The **two-phase simplex** — extends the simplex method to problems requiring
  artificial variables by first finding a basic feasible solution.
* The **primal-dual interior point** method — a log-barrier approach that moves
  through the interior of the feasible region and is competitive for large
  problems.

The package also includes utilities to convert problems to standard form,
check feasibility, compare methods, and visualise 2-D feasible regions with
**ggplot2**.

---

## 2. Setting Up an LP

The package ships with a built-in example dataset, `lp_example`, containing a
three-variable production LP.

```{r setup-lp}
data(lp_example)
str(lp_example)
```

`summarize_lp()` renders a human-readable description of any LP:

```{r summarize}
summarize_lp(
  obj      = lp_example$obj,
  A_ub     = lp_example$A_ub,
  b_ub     = lp_example$b_ub,
  maximize = lp_example$maximize
)
```

---

## 3. Converting to Standard Form

Before solving, every LP must be expressed in standard form:
$\min\ c^\top x$ subject to $Ax = b,\ x \geq 0$.
`lp_to_standard_form()` handles slack variables (for $\leq$ constraints),
surplus variables, and artificial variables automatically.

```{r standard-form}
sf <- lp_to_standard_form(
  obj      = lp_example$obj,
  A_ub     = lp_example$A_ub,
  b_ub     = lp_example$b_ub,
  maximize = lp_example$maximize
)

cat("Original variables :", sf$n_original, "\n")
cat("Slack variables    :", sf$n_slack,    "\n")
cat("Artificial vars    :", sf$n_artificial, "\n")
cat("Standard-form A dimensions:", nrow(sf$A), "x", ncol(sf$A), "\n")
sf$variable_names
```

The augmented constraint matrix `sf$A` now has identity sub-columns for each
slack variable, giving us a ready initial basis.

---

## 4. Solving with the Simplex Method

`simplex_method()` implements the full tableau simplex with Bland's rule
enabled by default to prevent cycling.

```{r simplex}
res_simplex <- simplex_method(
  obj      = lp_example$obj,
  A_ub     = lp_example$A_ub,
  b_ub     = lp_example$b_ub,
  maximize = lp_example$maximize
)

cat("Status     :", res_simplex$status, "\n")
cat("Objective  :", res_simplex$obj_value, "\n")
cat("Solution   :", res_simplex$solution, "\n")
cat("Iterations :", res_simplex$iterations, "\n")
cat("Runtime (s):", round(res_simplex$runtime, 4), "\n")
```

---

## 5. Solving with Two-Phase Simplex

The two-phase simplex method is required when the LP has equality constraints
or >= inequalities that prevent an obvious initial basis.  Phase 1 minimises
the sum of artificial variables; Phase 2 optimises the original objective.

```{r two-phase}
# Equality-constrained LP: min x1 + 2*x2  s.t.  x1 + x2 = 3
res_two <- two_phase_simplex(
  obj  = c(1, 2),
  A_eq = matrix(c(1, 1), nrow = 1),
  b_eq = 3
)

cat("Status            :", res_two$status, "\n")
cat("Objective value   :", res_two$obj_value, "\n")
cat("Solution          :", res_two$solution, "\n")
cat("Phase 1 iterations:", res_two$phase1_iterations, "\n")
cat("Phase 2 iterations:", res_two$phase2_iterations, "\n")
```

Infeasibility is automatically detected at the end of Phase 1:

```{r two-phase-infeasible}
# Contradictory equalities: x1+x2=5 AND x1+x2=3
res_infeas <- two_phase_simplex(
  obj  = c(1, 1),
  A_eq = matrix(c(1,1,1,1), nrow = 2, byrow = TRUE),
  b_eq = c(5, 3)
)
cat("Status:", res_infeas$status, "\n")
```

---

## 6. Solving with Interior Point

`interior_point()` implements the primal-dual log-barrier method.  Starting
from a strictly feasible interior point, it iteratively reduces the barrier
parameter $\mu$ until the duality gap falls below the tolerance.

```{r interior-point}
res_ip <- interior_point(
  obj      = lp_example$obj,
  A_ub     = lp_example$A_ub,
  b_ub     = lp_example$b_ub,
  maximize = lp_example$maximize
)

cat("Status     :", res_ip$status, "\n")
cat("Objective  :", round(res_ip$obj_value, 6), "\n")
cat("Solution   :", round(res_ip$solution, 6), "\n")
cat("Iterations :", res_ip$iterations, "\n")
cat("Final gap  :", round(tail(res_ip$duality_gap_history, 1), 8), "\n")
```

---

## 7. Comparing Methods

`compare_lp_methods()` runs all three solvers on the same problem and returns
both the full outputs and a tidy summary data frame.

```{r compare}
cmp <- compare_lp_methods(
  obj      = lp_example$obj,
  A_ub     = lp_example$A_ub,
  b_ub     = lp_example$b_ub,
  maximize = lp_example$maximize
)

knitr::kable(cmp$summary_df, digits = 6,
             caption = "Side-by-side comparison of LP solvers")
```

`plot_convergence()` visualises how each method converges:

```{r plot-convergence}
p_conv <- plot_convergence(cmp)
print(p_conv)
```

---

## 8. Visualising a 2-D LP

For problems with exactly two decision variables, `plot_lp_feasible_region()`
draws the feasible polygon, constraint boundaries, iso-objective contours, and
the optimal vertex using **ggplot2**.

```{r plot-region}
obj2  <- c(2, 3)
A_ub2 <- matrix(c(1, 1,
                   2, 1,
                   0, 1), nrow = 3, byrow = TRUE)
b_ub2 <- c(4, 6, 3)

p_region <- plot_lp_feasible_region(obj2, A_ub2, b_ub2,
                                     xlim = c(0, 7), ylim = c(0, 7))
print(p_region)
```

---

## 9. Advantages and Lessons Learned

Building **lpSolveR** provided hands-on exposure to a broad range of R
programming practices and mathematical concepts.

**Standard form conversion** is the foundation of all three solvers: by adding
slack, surplus, and artificial variables we reduce every LP to a canonical
$Ax = b,\ x \geq 0$ problem that the algorithms can handle uniformly.
**Simplex pivoting** then moves along the edges of the feasible polytope by
selecting an entering variable with a negative reduced cost and a leaving
variable via the minimum-ratio test; Bland's rule (smallest-index selection)
guarantees finite termination.  The **two-phase method** cleanly separates
feasibility (Phase 1) from optimality (Phase 2), and its artificial-variable
machinery makes it the natural choice when equality constraints are present.
**Barrier functions** (log-barrier) underpin the interior point solver: the
barrier term $-\mu \sum \ln x_i$ keeps iterates strictly feasible and shrinks
to zero as $\mu \to 0$, revealing the true optimum; the duality gap $x^\top s
/ n$ serves as a rigorous convergence certificate rooted in **duality theory**.

On the software side, **roxygen2 documentation** was used throughout to
generate `man/` pages and the NAMESPACE file, keeping documentation
co-located with the code.  **testthat unit testing** verified correctness on
known problems and edge cases (infeasibility, equality constraints), making
refactoring safe.  The **devtools workflow** (`document()`, `check()`,
`build()`) streamlined the package development cycle.  **ggplot2
visualisation** powered both the feasible-region plot and the convergence plot,
with `aes()` and `geom_*` layers composited programmatically.  Finally, **data
packaging with `usethis`/`save()`** embedded the `lp_example` dataset directly
in the package, making examples fully reproducible without external file paths.
Together these tools demonstrate that a self-contained, documented, and tested
R package can be built entirely in base R plus a handful of well-maintained
CRAN packages.
